The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages

### 2024-10-29 submission/manuscript.Rmd
new intro draft

A pivotal step in the evolution of multicellularity is the division labor among cellular lineages, including the the distinction of the germline from the soma. In the early embryo, genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications. Failure to repress germline genes in somatic lineages is common signature of cancer and observed in select neurodevelopmental disorders, however it is currently unclear how factors like development and sex influence their repression and somatic misexpression. Here, we examine how cellular context influences the development of somatic tissue identity in mice with loss of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, ovary, and testis genes within the brain. Late-stage spermatogenesis genes, but not somatic testicular genes, were highly enriched within the _Kdm5c_-KO brain, indicating an erosion of soma-germline boundary. Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation. However, late-stage spermatogenesis genes devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.

@@

- While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in late oogenesis and spermatogenesis, employ the same silencing mechanisms and how silencing changes over the course of development. 
- In mammals, genes crucial for early germline specification gain repressive DNA and histone modifications in the early embryo. 
- Germline genes have not been comprehensively assessed as a whole

@@

A single genome holds the instructions to generate the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs. Loss of chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.

To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. Here we focus on the X chromosome gene lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes, given the lack of a comprehensive list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for different classes of germline genes, e.g. meiotic versus spermatid differentiation genes.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated a list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time. Additionally, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants to illuminate the impact of sex upon germline gene suppression. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

>>>>

A pivotal step in the evolution of multicellularity is the division labor among cellular lineages, including distinguishing the germline from the soma. In the early embryo, genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications. Failure to repress germline genes in somatic lineages is common signature of cancer and observed in select neurodevelopmental disorders, however it is currently unclear how factors like development and sex influence their repression and somatic misexpression. Here, we examine how cellular context influences the development of somatic tissue identity in mice with loss of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, ovary, and testis genes within the brain. By developing a comprehensive list of mouse germline-enriched genes, we found late-stage spermatogenesis genes and not somatic testicular genes were highly enriched within the _Kdm5c_-KO brain. Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression. Although many germline mRNAs are expressed in the _Kdm5c_-KO brain and EpiLCs, KDM5C recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation. Late-stage spermatogenesis genes lacking KDM5C and devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.

@@

- In mammals, genes crucial for early germline specification gain repressive DNA and histone modifications in the early embryo. 
- Germline genes have not been comprehensively assessed as a whole

@@

A single genome holds the instructions for generating the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed impaired chromatin regulation commonly occurs in both cancer[@zhaoLanguageChromatinModification2021; @chiCovalentHistoneModificationsmiswritten2010; @berdascoAberrantEpigeneticLandscape2010] and neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. While many studies have identified their importance for regulating tumor suppressor genes and brain-specific transcriptional programs, loss of chromatin regulators can also cause ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of testis genes in colon tumors[@feichtingerMetaanalysisClinicalData2012] or liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms driving ectopic gene expression and its impact upon cancer and neurodevelopment are still poorly understood.

Separation of germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[XXX, @devlinPanmetazoanViewGermlinesoma2023]. In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in later stages of oogenesis and spermatogenesis, employ the same silencing mechanisms. Furthermore, because many studies have focused on the silencing of key marker genes during early male embryogenesis, much is unknown about how cellular context influences the manifestation of germline gene misexpression. Given that impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[XXX; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021], elucidating the impact of factors like sex and development is crucial for determining the mechanisms governing long-term misexpression of germline genes.

Here, we employed a genome-wide analyses to explore the impact of cellular context upon loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C). KDM5C lies on the X chromosome and erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Somatic loss of KDM5C promotes tumorigenicity in a variety of cancer types[@shenSuppressionEnhancerOveractivation2016; @chenE6ProteinExpressed2018; @zhengDeficiencyXinactivationEscaping2021], while germline mutations cause the NDD Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some testis germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, what mechanisms underlie their dysregulation, and how KDM5C interacts with other known germline silencing mechanisms.

To illuminate KDM5C's role in tissue identity, we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated a list of mouse germline-enriched genes, enabling genome-wide analysis of germline gene silencing mechanisms for the first time. Additionally, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants to illuminate the impact of sex upon germline gene suppression. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
